Arrows for Parallel Computations

### 2017-12-16 paper/content/topologySkeletons.tex

If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing. This materializes with the usage of |loop| of the |ArrowLoop| type class and is probably due to the way their respective parallelism engines work internally. \mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| type class. This allows us for selective enabling and disabling of parallelism.
\begin{code}
class ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
\end{code}
As Eden has no problems with the looping skeletons, we declare a delegating instance:
\begin{code}
instance ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN = parEvalN
\end{code}
The Par Monad and GpH backends have parallelism disabled in their instance of |FutureEval|. This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
\begin{code}
instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
    ArrowParallel arr a b conf) => FutureEval arr a b conf where
    evalN _ = listApp
\end{code}

@@

@@

\begin{figure}[t]
\begin{code}
pipeSimple :: (ArrowLoop arr, FutureEval arr a a conf) =>
	conf -> [arr a a] -> arr a a
pipeSimple conf fs =
	loop (arr snd &&&
		(arr (uncurry (:) >>> lazy) >>> evalN conf fs)) >>>
	arr last
\end{code}
\caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows. Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input |[a]| terminates before passing it into |evalN|.}
\label{fig:pipeSimple}

\begin{code}
pipe :: (ArrowLoop arr, FutureEval arr (fut a) (fut a) conf, Future fut a) =>
	conf -> [arr a a] -> arr a a
pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
\end{code}
\caption{Final definition of the |pipe| skeleton with Futures.}
\label{fig:pipe}
\end{figure}

Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. |arr a b| and |arr b c|. By wrapping these two arrows inside a bigger arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
\begin{figure}[tb]
\begin{code}
pipe2 :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
         FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g =
	(arr return &&& arr (const [])) &&& arr (const []) >>>
	pipe conf (replicate 2 (unify f g)) >>>
	arr snd >>> arr head where
			unify :: (ArrowChoice arr) => arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
			unify f g =
				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
				arr (\((a, b), c) -> ((c, a), b))

(parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
	FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
	arr a b -> arr b c -> arr a c
(parcomp) = pipe2 ()
\end{code}
\caption{Definition of |pipe2| and a parallel |>>>|.}
\label{fig:pipe2}
\end{figure}

@@

Another version of |>>>| is:
\begin{code}
f parcomp g = (f . put) >>> (get . g)
\end{code}
It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them. Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.

>>>>

If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing. This materializes with the usage of |loop| of the |ArrowLoop| type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden. An investigation of where this difference comes from is beyond the scope of this work, we only provide a workaround for these types of skeletons as such skeletons probably are not of much importance outside of a distributed memory environment. However our workaround enables users of the DSL to test their code within a shared memory setting even though other skeletons would be better suited to be run with them.
The idea of the fix is to provide a |ArrowLoopParallel| typeclass that has two functions - |loopParEvalN| and |postLoopParEvalN| where the first is to be used inside an |loop| construct while the latter will be used right outside of the |loop|. This way we can delegate to the actual |parEvalN| in the spot where the backend supports it.
\begin{code}
class ArrowParallel arr a b conf => ArrowLoopParallel arr a b conf where
    loopParEvalN :: conf -> [arr a b] -> arr [a] [b]
    postLoopParEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{code}
As Eden has no problems with the looping skeletons, we use this instance:
\begin{code}
instance (ArrowChoice arr, ArrowParallel arr a b Conf) => ArrowLoopParallel arr a b Conf where
    loopParEvalN = parEvalN
    postLoopParEvalN _ = evalN
\end{code}
As the backends for the |Par| Monad and GpH have problems with |parEvalN| inside of |loop| their respective instances for |ArrowLoopParallel| look like this:
\begin{code}
instance (ArrowChoice arr, ArrowParallel arr a b (Conf b)) => ArrowLoopParallel arr a b (Conf b) where
    loopParEvalN _ = evalN
    postLoopParEvalN = parEvalN
\end{code}

@@

\begin{figure}[t]
\begin{code}
pipeSimple :: (ArrowLoop arr, ArrowLoopParallel arr a a conf) =>
	conf -> [arr a a] -> arr a a
pipeSimple conf fs =
    loop (arr snd &&&
        (arr (uncurry (:) >>> lazy) >>> loopParEvalN conf fs)) >>>
    arr last
\end{code}
\caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows. Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input |[a]| terminates before passing it into |evalN|.}
\label{fig:pipeSimple}

@@

\begin{code}
pipe :: (ArrowLoop arr, ArrowLoopParallel arr (fut a) (fut a) conf,
	Future fut a conf) =>
	conf -> [arr a a] -> arr a a
pipe conf fs = unliftFut conf (pipeSimple conf (map (liftFut conf) fs))

liftFut :: (Arrow arr, Future fut a conf, Future fut b conf) =>
	conf -> arr a b -> arr (fut a) (fut b)
liftFut conf f = get conf >>> f >>> put conf

unliftFut :: (Arrow arr, Future fut a conf, Future fut b conf) =>
	conf -> arr (fut a) (fut b) -> arr a b
unliftFut conf f = put conf >>> f >>> get conf
\end{code}
\caption{Final definition of the |pipe| skeleton with Futures.}
\label{fig:pipe}
\end{figure}

Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. |arr a b| and |arr b c|. By wrapping these two arrows inside a bigger arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
\begin{figure}[tb]
\begin{code}
pipe2 :: (ArrowLoop arr, ArrowChoice arr,
    ArrowLoopParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
    Future fut (([a], [b]), [c]) conf) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g =
	(arr return &&& arr (const [])) &&& arr (const []) >>>
    pipe conf (replicate 2 (unify f g)) >>>
    arr snd >>>
    arr head
    where
        unify :: (ArrowChoice arr) => 
			arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
        unify f' g' =
			(mapArr f' *** mapArr g') *** arr (const []) >>>
			arr (\((b, c), a) -> ((a, b), c))

(parcomp) :: (ArrowLoop arr, ArrowChoice arr,
    ArrowLoopParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
    Future fut (([a], [b]), [c]) ()) =>
    arr a b -> arr b c -> arr a c
(parcomp) = pipe2 ()
\end{code}
\caption{Definition of |pipe2| and |(parcomp)|, a parallel |>>>|.}
\label{fig:pipe2}
\end{figure}

@@

%Another version of |>>>| is:
%\begin{code}
%f parcomp g = (f . put) >>> (get . g)
%\end{code}
%It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them. Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.
